Escape Analysis and Specialization in a Tracing JIT

### 2010-09-26 talk/pepm2011/paper.tex
whack at the introduction

The goal of a just-in-time compiler for a dynamic language is obviously to
improve the speed of the language over an implementation of the language that
uses interpretation. The first goal of a JIT is thus to remove the
interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
overhead of the interpreter's data structures, such as operand stack etc. The
second important problem that any JIT for a dynamic language needs to solve is
how to deal with the overhead of boxing of primitive types and of type
dispatching. Those are problems that are usually not present in statically typed
languages.

Boxing of primitive types means that dynamic languages need to be able to handle
all objects, even integers, floats, etc. in the same way as user-defined
instances. Thus those primitive types are usually \emph{boxed}, i.e. a small
heap-structure is allocated for them, that contains the actual value.

Type dispatching is the process of finding the concrete implementation that is
applicable to the objects at hand when doing a generic operation on them. An
example would be the addition of two objects: The addition needs to check what
the concrete objects that should be added are, and choose the implementation
that is fitting for them.

Last year, we wrote a paper \cite{XXX} about how PyPy's meta-JIT
approach works. These explain how the meta-tracing JIT can remove the overhead
of bytecode dispatch. In this paper we want to explain how the traces that are
produced by our meta-tracing JIT are then optimized to also remove some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. The most important technique to achieve this is a form of
escape analysis \cite{XXX} that we call \emph{virtual objects}. This is best
explained via an example.

@@

\section{Escape Analysis in a Tracing JIT}
\label{sec:Escape Analysis in a Tracing JIT}

>>>>

The goal of a just-in-time compiler for a dynamic language is obviously to
improve the speed of the language over an implementation of the language that
uses interpretation. The first goal of a JIT is thus to remove the
interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
overhead of the interpreter's data structures, such as operand stack etc. The
second important problem that any JIT for a dynamic language needs to solve is
how to deal with the overhead of boxing of primitive types and of type
dispatching. Those are problems that are usually not present or at least less
severe in statically typed languages.

Boxing of primitive types means that dynamic languages need to be able to handle
all objects, even integers, floats, bools etc. in the same way as user-defined
instances. Thus those primitive types are usually \emph{boxed}, i.e. a small
heap-structure is allocated for them, that contains the actual value. Boxing
primitive types can be very costly, because XXX

Type dispatching is the process of finding the concrete implementation that is
applicable to the objects at hand when doing a generic operation on them. An
example would be the addition of two objects: The addition needs to check what
the concrete objects that should be added are, and choose the implementation
that is fitting for them. Type dispatching is a very common operation in a
dynamic language because no types are known at compile time, so all operations
need it.

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT often takes the form of an
extension to an existing interpreter, which can be sped up that way. The PyPy
project is an environment for implementing dynamic programming languages. It's
approach to doing so is to straightforwardly implement an interpreter for the
to-be-implemented language, and then use powerful tools to turn the interpreter
into an efficient VM that also contains a just-in-time compiler. This compiler
is automatically generated from the interpreter using partial-evaluation-like
techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
tracing JIT compilers is described in Section~\ref{sec:Background}

The tracing JIT approach that the PyPy project is taking removes the overhead
of bytecode dispatch. In this paper we want to explain how the traces that are
produced by our meta-tracing JIT can be optimized to also remove some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. To understand the problem more closely, we analyze the
occurring object lifetimes in Section~\ref{sec:lifetimes}. The most important
technique to achieve this is a form of escape analysis \cite{XXX} that we call
\emph{virtual objects}, which is described in Section~\ref{sec:virtuals}. The
basic approach of virtual objects can then be extended to also be used for
type-specializing the traces that are produced by the tracing JIT
(Section~\ref{sec:crossloop}). In Section~\ref{sec:XXX} we describe some
supporting techniques that are not central to the approach, but are needed to
improve the results. The introduced techniques are evaluated in
Section~\ref{sec:Evaluation}.

The contributions of this paper are:

\begin{enumerate}
    \item An efficient and effective escape analysis for a tracing JIT
    \item XXX
\end{enumerate}

@@

\section{Escape Analysis in a Tracing JIT}
\label{sec:virtuals}
